---
title: "Final"
format: html
author: Rylee sheils 
---

GitHub repository: [ENVS-193DS_winter-2026_final](https://github.com/ryleesheils/ENVS-193DS_winter-2026_final)

# Set Up 
```{r}
#| label: Set Up 
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
library(forcats)
library(here)
```

# Problem 1. Research writing

## a. Transparent Statisitcal Methods 
In part 1, they used a simple linear regression because they tested whether precipitation predicts flooded wetland area. In part 2, they used a Kruskal–Wallis rank-sum test because they tested whether median flooded wetland area differs across five water year classification groups.

## b. Figure Needed
A good figure for part 2 would be a box plot with water year classification on the x-axis and flooded wetland area on the y-axis. I think this would be the best choice because it makes it easy to compare the median flooded wetland area across all five groups and also shows how spread out the data are in each one. I would also add the individual data points so it is easier to see the variation and overlap between groups.

## c. Suggestions for Rewriting 
Precipitation did not significantly predict flooded wetland area in the San Joaquin River Delta. This result was based on a simple linear regression (test statistic = not provided, distribution = not provided, degrees of freedom = not provided, p = 0.11, α = significance level not provided).

Median flooded wetland area also did not differ significantly across water year classifications. This result was based on a Kruskal-Wallis rank-sum test comparing wet, above normal, below normal, dry, and critical drought years (test statistic = not provided, distribution = chi-squared, degrees of freedom = 4, p = 0.12, α = significance level not provided).

## d. Interpretation
One additional variable that could influence flooded wetland area is median daily Delta outflow during the spring management period. This would be a continuous variable, and it could affect flooded wetland area because the paper found flooded wetland area was weakly correlated with Delta outflow during a key time for wetland water management.

# Problem 2. Data Visualization 

## a. Cleaning and Summarizing 

```{r}
#| message: false
#| warning: false
kelp <- read_csv(here("data", "kelp_data.csv"))
```

```{r}
#| message: false
#| warning: false
kelp_clean <- kelp |>
  clean_names() |>
  # cleans the column names
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |>
  # keeps only the three sites of interest
  
  mutate(site_name = case_when(
    site == "NAPL" ~ "Naples",
    site == "IVEE" ~ "Isla Vista",
    site == "CARP" ~ "Carpinteria"
  )) |>
  # creates a new column with full site names
  
  mutate(site_name = as_factor(site_name),
         site_name = fct_relevel(site_name, "Naples", "Isla Vista", "Carpinteria")) |>
  # makes site_name a factor and sets the order
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>
  # removes the flagged incorrect observation
  
  group_by(site_name, year) |>
  # groups the data by site and year
  
  summarize(mean_fronds = mean(fronds)) |>
  # calculates the mean fronds for each site-year combination
  
  ungroup()
  # removes grouping
```

```{r}
slice_sample(kelp_clean, n = 5)
```

```{r}
str(kelp_clean)
```

## b. Understanding the Data 
The observation from 2000-09-22 in transect 6 and quad 40 is filtered out because the `fronds` value is `-99999`, which is a missing-data code rather than a real measurement. In this dataset, `FRONDS` represents the number of giant kelp fronds greater than 1 meter tall, and for this row the value appears because the datasheet was missing, so it should not be treated as an actual frond count. 

## c. Visualize the Data
```{r}
ggplot(kelp_clean, aes(x = year, y = mean_fronds, color = site_name)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(
    name = "Site",
    values = c(
      "Naples" = "#157A76",
      "Isla Vista" = "#8E7767",
      "Carpinteria" = "#5E776F"
    )
  ) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  labs(
    title = "Sites vary in mean kelp fronds per year",
    x = "Year",
    y = "Mean kelp fronds per year"
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.border = element_blank(),
    legend.position = "right"
  )
```

## d. 
The figure shows that mean kelp fronds vary across both year and site. Isla Vista tends to have the highest values and the biggest changes over time, while Naples and Carpinteria are usually lower. Overall, the plot suggests that kelp abundance is not constant and differs depending on location and year.

# Problem 3. Data Analysis

## Set up 
```{r}
mopl <- read_csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv"))
```

## a. Response Varirables 
In this data set, a value of 1 means the location was a nest site that was used by a Mountain Plover, and a value of 0 means the location was an available site that was not used for nesting. Biologically, the response variable represents nest-site use, or whether a Mountain Plover selected that location for nesting.

## b. 
`vo_nest` is a continuous variable because it is measured numerically. As visual obstruction increases, the probability of Mountain Plover nest-site use would probably decrease because plovers tend to choose open nesting habitat with shorter and less dense vegetation. I found this in the Introduction where the paper says Mountain Plovers typically breed in areas with extensive bare ground and short, sparse vegetation, and in the Nest-Site Selection methods where visual obstruction was included as a habitat variable.

`edgedist` is a continuous variable because it measures distance to the prairie dog colony edge. As distance to colony edge increases, the probability of nest-site use would probably decrease because the paper suggests plovers are more likely to use areas closer to colony edges than deep colony cores. I found this in the Discussion where the authors say probability of site use declined near the cores of large colonies and that distance to colony edge was a strong driver of plover presence. 