---
title: "Final"
format: html
author: Rylee sheils 
---

GitHub repository: [ENVS-193DS_winter-2026_final](https://github.com/ryleesheils/ENVS-193DS_winter-2026_final)

# Set Up 
```{r}
#| label: Set Up
#| message: false
#| warning: false
library(tidyverse) #load in tidyverse
library(janitor) #load in janitor
library(forcats) #load in forcats
library(here) #load in here
library(DHARMa) #load in DHARMa

```

# Problem 1. Research writing

## a. Transparent Statisitcal Methods 
In part 1, they used a simple linear regression because they tested whether precipitation predicts flooded wetland area. In part 2, they used a Kruskal–Wallis rank-sum test because they tested whether median flooded wetland area differs across five water year classification groups.

## b. Figure Needed
A good figure for part 2 would be a box plot with water year classification on the x-axis and flooded wetland area on the y-axis. I think this would be the best choice because it makes it easy to compare the median flooded wetland area across the five groups and also shows how spread out the data are in each one. I would also add the individual data points so it is easier to see the variation and overlap between groups.

## c. Suggestions for Rewriting 
Precipitation did not significantly predict flooded wetland area in the San Joaquin River Delta. This result was based on a simple linear regression (test statistic = not provided, distribution = not provided, degrees of freedom = not provided, p = 0.11, α = significance level not provided).

Median flooded wetland area also did not differ significantly across water year classifications. This result was based on a Kruskal-Wallis rank-sum test comparing wet, above normal, below normal, dry, and critical drought years (test statistic = not provided, distribution = chi-squared, degrees of freedom = 4, p = 0.12, α = significance level not provided).

## d. Interpretation
One additional variable that could influence flooded wetland area is median daily outflow during the spring management period. This would be a continuous variable, and it could affect flooded wetland area as the paper found flooded wetland area was weakly correlated with Delta outflow during a key time for wetland water management.

# Problem 2. Data Visualization 

## a. Cleaning and Summarizing 

Read in Kelp Data
```{r}
#| label: Read in Kelp Data
#| message: false
#| warning: false
kelp <- read_csv(here("data", "kelp_data.csv")) #Read in Kelp data and make data frame called kelp
```

```{r}
#| message: false
#| warning: false
kelp_clean <- kelp |> # make new data frame called data clean 
  clean_names() |> # cleans the column names
  
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # keeps only the three sites of interest
  
  mutate(site_name = case_when(
    site == "NAPL" ~ "Naples", # when code say NAPL change to naples
    site == "IVEE" ~ "Isla Vista", #when code says IVEE change to Isla Vista 
    site == "CARP" ~ "Carpinteria" # when code say CARP change to Carpenteria 
  )) |> # creates a new column with full site names
  
  mutate(site_name = as_factor(site_name), #change the site names 
         site_name = fct_relevel(site_name, "Naples", "Isla Vista", "Carpinteria")) |>  # makes site_name a factor and sets the order
  
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> #filter data 
  # removes the flagged incorrect observation
  
  group_by(site_name, year) |># groups the data by site and year
  
  summarize(mean_fronds = mean(fronds)) |> # calculates the mean for each site-year combination
  
  ungroup() # removes groups
```

```{r}
slice_sample(kelp_clean, n = 5) #show sample of clean data
```

```{r}
str(kelp_clean) #show str 
```

## b. Understanding the Data 
The observation from 2000-09-22 in transect 6 and quad 40 is filtered out because the `fronds` value is `-99999`, which is a missing-data code rather than a real measurement. In this dataset, `FRONDS` represents the number of giant kelp fronds greater than 1 meter tall, and for this row the value appears because the datasheet was missing, so it should not be treated as an actual frond count. 

## c. Visualize the Data
```{r}
ggplot(kelp_clean, aes(x = year, y = mean_fronds, color = site_name)) + #create plot with clean data and add x and y variables 
  geom_line(linewidth = 0.7) + #change width of line 
  geom_point(size = 2) + #change size of points
  scale_color_manual( #colors of different varaibles
    name = "Site", 
    values = c(
      "Naples" = "#157A76", #color for naples
      "Isla Vista" = "#8E7767", #color for isla vista
      "Carpinteria" = "#5E776F" #color for carpinteria 
    )
  ) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) + #make scales 
  labs(
    title = "Sites vary in mean kelp fronds per year", #label title 
    x = "Year", #x axis label
    y = "Mean kelp fronds per year" # y axis label 
  ) +
  theme_minimal(base_family = "sans") + #set theme
  theme(
    plot.title = element_text(hjust = 0.5), #change plot elements 
    panel.border = element_blank(),
    legend.position = "right" #change position of legend 
  )
```

## d. 
The figure shows that mean kelp fronds vary across both year and site. Isla Vista tends to have the highest values and the biggest changes over time, while Naples and Carpinteria are usually lower. Overall, the plot suggests that kelp abundance is not constant and differs depending on location and year.

# Problem 3. Data Analysis


## Set up 
```{r}
#| message: false
#| warning: false
mopl <- read_csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv"))
```

## a. Response Varirables 
In this data set, a value of 1 means the location was a nest site that was used by a Mountain Plover, and a value of 0 means the location was an available site that was not used for nesting. Biologically, the response variable represents nest-site use, or whether a Mountain Plover selected that location for nesting.

## b. 
`vo_nest` is a continuous variable because it is measured numerically. As visual obstruction increases, the probability of Mountain Plover nest-site use would probably decrease because plovers tend to choose open nesting habitat with shorter and less dense vegetation. I found this in the Introduction where the paper says Mountain Plovers typically breed in areas with extensive bare ground and short, sparse vegetation, and in the Nest-Site Selection methods where visual obstruction was included as a habitat variable.

`edgedist` is a continuous variable because it measures distance to the prairie dog colony edge. As distance to colony edge increases, the probability of nest-site use would probably decrease because the paper suggests plovers are more likely to use areas closer to colony edges than deep colony cores. I found this in the Discussion where the authors say probability of site use declined near the cores of large colonies and that distance to colony edge was a strong driver of plover presence. 

## c. 
| Model number | Visual obstruction | Distance to prairie dog colony | Predictor list |
|---|---|---|---|
| 1 |  |  | no predictors (null model) |
| 2 | x |  | visual obstruction |
| 3 |  | x | distance to prairie dog colony edge |
| 4 | x | x | all predictors (full model) |

## d. 
```{r}
mopl_clean <- mopl |>
  clean_names() |>
  mutate(
    used_bin = case_when(
      id == "used" ~ 1,
      id == "available" ~ 0
    )
  ) |>
  select(used_bin, edgedist, vo_nest) |>
  drop_na()
```

```{r}
# model 1: null model
model1 <- glm(
  used_bin ~ 1,
  data = mopl_clean,
  family = binomial
)

# model 2: visual obstruction only
model2 <- glm(
  used_bin ~ vo_nest,
  data = mopl_clean,
  family = binomial
)

# model 3: distance to colony edge only
model3 <- glm(
  used_bin ~ edgedist,
  data = mopl_clean,
  family = binomial
)

# model 4: full model
model4 <- glm(
  used_bin ~ vo_nest + edgedist,
  data = mopl_clean,
  family = binomial
)
```

## e. 

```{r}
AIC(model1, model2, model3, model4)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the model predicting Mountain Plover nest-site use from visual obstruction. This model had the lowest AIC value (331.769), so it provided the best fit among the candidate models while balancing model complexity.

## f. 

```{r}

model2_sim <- simulateResiduals(fittedModel = model2)

plot(model2_sim)
```
I checked the diagnostics for the selected logistic regression model using simulated DHARMa residuals. The diagnostic plots did not show any significant problems, and the formal tests suggested no evidence of poor residual fit, overdispersion, or unusual outliers. Overall, the model appears to meet the assumptions reasonably well.

From the figure:

KS test p = 0.72382 → no evidence the residual distribution is off

Dispersion test p = 0.992 → no evidence of overdispersion

Outlier test p = 1 → no evidence of problematic outliers

the top-right panel also says “No significant problems detected"

## g. 
```{r}
pred_data <- tibble(
  vo_nest = seq(
    from = min(mopl_clean$vo_nest, na.rm = TRUE),
    to = max(mopl_clean$vo_nest, na.rm = TRUE),
    length.out = 100
  )
)

predictions <- predict(model2, newdata = pred_data, type = "link", se.fit = TRUE)

pred_data <- pred_data |>
  mutate(
    fit_link = predictions$fit,
    se_link = predictions$se.fit,
    fit_prob = plogis(fit_link),
    lower_ci = plogis(fit_link - 1.96 * se_link),
    upper_ci = plogis(fit_link + 1.96 * se_link)
  )
```

```{r}
ggplot() +
  geom_jitter(
    data = mopl_clean,
    aes(x = vo_nest, y = used_bin),
    height = 0.03,
    width = 0,
    alpha = 0.4,
    color = "#5B8A72"
  ) +
  geom_ribbon(
    data = pred_data,
    aes(x = vo_nest, ymin = lower_ci, ymax = upper_ci),
    fill = "#A5C4B0",
    alpha = 0.4
  ) +
  geom_line(
    data = pred_data,
    aes(x = vo_nest, y = fit_prob),
    color = "#1F5A4A",
    linewidth = 1.2
  ) +
  labs(
    x = "Visual obstruction at nest",
    y = "Predicted probability of Mountain Plover nest-site use"
  ) +
  theme_classic()
```

## h. 
Figure 1. Predicted probability of Mountain Plover nest-site use as a function of visual obstruction at the nest. Points show the underlying binary nest-site use data (0 = available but unused site, 1 = used nest site), the solid line shows predicted probabilities from the selected logistic regression model, and the shaded band shows 95% confidence interval around the predictions. Data source: Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance [Dataset]. Dryad.

## i. 
```{r}
predict(
  model2,
  newdata = tibble(vo_nest = 0),
  type = "response"
)
```
At a visual obstruction value of 0, the predicted probability of Mountain Plover nest-site occupancy is 0.732. This means the model predicts about a 73.2% probability that a site with no visual obstruction would be used as a nest site.

## j. 
The results suggest that visual obstruction is an important predictor of Mountain Plover nest-site use. As visual obstruction increases, the predicted probability of nest-site use decreases, which suggests that Mountain Plovers are more likely to select open nesting areas with little visual blockage. This matches the ecological expectations from the paper, which describes Mountain Plovers as birds that prefer relatively flat areas with bare ground and sparse vegetation.

# Problem 4. Affective and Exploratory Visualizations

## a. Comparing Visualizations 

### How are the visualizations different?
My Homework 2 exploratory visualizations represented the data in a more standard statistical way, using separate plots to isolate one relationship at a time, like cloud cover and energy or temperature and energy. By contrast, my affective visualization turns the same daily observations into a hand-drawn flower garden, where each flower stands for one day and multiple variables are layered into the same symbol through stem height, petal number, color, leaves, and flower center.

### Similarities 
All of my visualizations are still based on the same personal daily data and focus on understanding what shapes my energy over time. Even though the style changes, they all keep energy as the main outcome and try to connect it to other daily conditions like cloud cover, sleep, and other factors that may influence how I feel.

### Patterns 
Across both the exploratory and affective visualizations, I see that energy seems to shift with weather and other daily conditions rather than staying constant. In Homework 2, I wrote that higher temperatures seemed to be associated with slightly higher energy levels and that cloud cover might also matter, while the flower visualization reinforces that energy is highest on some bright, open-feeling days but also shows that sleep, assignments, caffeine, and steps are happening at the same time. The patterns are not completely different, but the affective visualization makes the data feel more layered and personal because several variables are shown together. Something I did notice is that I tended to have very low energy levels on Friday's which is interesting as that is my final day of the week for assignments and school. 

### Feedback 
During week 9 workshop my group gave me the fantastic idea of visualizing my data through flowers. I implemented that process by making sure each flower feature had a clear meaning and that the final piece had a readable legend, consistent visual structure, and a clear connection between the timeline and the variables being represented. After presenting my visualization in week 10, it was clear that this was a solid choice as my peers said it was easy to follow yet displayed a lot of different variables of data. 

## b. Designing Analysis

Research question:
Do energy levels differ across cloud cover conditions?

I would use a Kruskal-Wallis test to answer this question. My response variable is energy level measured on a 1 to 5 scale, and my predictor variable is cloud cover grouped into the categories clear, cloudy, and rain. This analysis is appropriate because I am comparing energy across more than two categorical weather groups, and Kruskal-Wallis is the non-parametric test used for this type of comparison.

## c. Check Assumptions and Run Analysis 
Because my response variable was energy level on a 1 to 5 scale and my predictor variable was cloud cover category, I used a non-parametric Kruskal-Wallis test instead of a parametric ANOVA. I first checked the number of observations in each recoded cloud cover group and looked at the underlying data with a jitter plot to make sure the categories were reasonable for comparison.
```{r}
daily_data <- read_csv(here("data", "ENVSdsresearch.csv"))
```

```{r}
daily_clean <- daily_data |>
  clean_names() |>
  mutate(
    cloud_group = case_when(
      cloud_cover == "mostly clear" ~ "clear",
      cloud_cover == "partly cloudy" ~ "cloudy",
      cloud_cover == "cloudy" ~ "cloudy",
      cloud_cover == "cloudy/rain" ~ "rain"
    ),
    cloud_group = as_factor(cloud_group),
    cloud_group = fct_relevel(cloud_group, "clear", "cloudy", "rain")
  ) |>
  select(cloud_group, energy_level_1_5)

```

```{r}
count(daily_clean, cloud_group)
```

```{r}
ggplot(daily_clean, aes(x = cloud_group, y = energy_level_1_5, color = cloud_group)) +
  geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.7) +
  labs(
    x = "Cloud cover condition",
    y = "Energy level (1–5)"
  ) +
  theme_classic()
```

```{r}
kruskal.test(energy_level_1_5 ~ cloud_group, data = daily_clean)
```

```{r}
kruskal_result <- kruskal.test(energy_level_1_5 ~ cloud_group, data = daily_clean)

n <- nrow(daily_clean)
k <- nlevels(daily_clean$cloud_group)
H <- as.numeric(kruskal_result$statistic)

eta_squared_H <- (H - k + 1) / (n - k)

eta_squared_H
```

I ran a Kruskal-Wallis test to compare energy levels across cloud cover conditions. I used the recoded categories clear, cloudy, and rain, where clear represented mostly clear days, cloudy combined partly cloudy and cloudy days, and rain represented cloudy/rain days. The test was not statistically significant, 
𝜒
2
=
3.475
χ
2
=3.475, 
𝑑
𝑓
=
2
df=2, 
𝑝
=
0.176
p=0.176, which suggests there is not strong evidence that typical energy level differed across cloud cover groups in this sample. The effect size was small (
𝜂
2
≈
0.055
η
2
≈0.055), suggesting cloud cover explained only a small amount of the variation in energy levels.

## d. Create Visualization 

```{r}
ggplot(daily_clean, aes(x = cloud_group, y = energy_level_1_5, color = cloud_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2, width = 0.5) +
  geom_jitter(width = 0.15, height = 0, size = 2.4, alpha = 0.8) +
  scale_color_manual(values = c(
    "clear" = "#F2C94C",
    "cloudy" = "#56B4E9",
    "rain" = "#8E63CE"
  )) +
  labs(
    x = "Cloud cover condition",
    y = "Energy level (1–5)"
  ) +
  theme_classic()
```
To reflect the Kruskal-Wallis analysis, I created a boxplot with jittered points showing energy level across recoded cloud cover groups. This visualization makes it easier to compare the spread and typical values of energy across the weather categories while still showing the underlying observations.

## e. Write a Caption 
Figure 2. Energy levels across recoded cloud cover conditions. Points show individual daily observations of energy level on a 1 to 5 scale, while boxplots summarize the distribution of energy within the clear, cloudy, and rain groups. Colors were chosen to match the weather categories used in the affective flower visualization.

## f. write about your results 
The visualization and Kruskal-Wallis test both suggest that cloud cover may relate to energy level, but the evidence was not strong enough in this sample to conclude that energy differed significantly across the cloud cover groups. The effect size was small, which suggests cloud cover explained only a small amount of the variation in my energy. This fits with my affective visualization, which shows that energy is likely shaped by several daily factors at once rather than weather alone.

## g. Week 10 Attendance 
I was there! 



